Spitzer, Gaia and the Potential of the Milky Way

### 2013-06-22 docs/paper1/paper.tex
sketch out structure, add Gaia info

% Above here is from KVJ. Below is APW.

\section{Method}
With access to full phase-space information for stars in an associated
debris structure, it is no longer necessary to model the full orbit of
the stream. Instead, it is possible to use the individual stars as
potential measures by exploting the simple fact that the stars must
have come from the same progenitor. If the present-day orbital
parameters of the progenitor system are known, this places an
intuitive constraint on the potential of the Milky Way: if the orbits
of the stars and the orbit of the progenitor are integrated
\emph{backwards} in time, the stars should `recombine' with the
progenitor. Conversely, if the potential is incorrect, the orbits of
the stars will diverge from that of the progenitor and thus will not
be recaptured by the satellite system.

The starting point for applying this method is a sample of stars with
known full space kinematics $(\bs{x}_i, \bs{v}_i)|_{t=0}$, the orbital
parameters for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$,
and a functional form for the potential, $\Phi$\footnote{Future work
  will investigate a more general ... basis function blah blah.}. For
a given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for some time (determined from an
estimate of the interaction time of the system). [APW: explain that
  this is not Nbody, just test particles!] At each timestep $t_j$, for
each particle $i$, a set of normalized, relative phase-space
coordinates, $(\bs{q}_i(t_j), \bs{p}_i(t_j))$, are computed using the
instantaneous tidal radius, escape velocity, and phase-space position
of the progenitor. At a particular timestep, these coordinates are
defined as
\begin{equation}
  \bs{q}_i = \frac{\bs{x}_i -
    \bs{x}_p}{R_{tide}}\,\,\,\,\,\,\,\,\bs{p}_i = \frac{\bs{v}_i -
    \bs{v}_p}{v_{esc}}.
\end{equation}
These definitions require an estimate of the mass of the satellite,
$m_{sat}$, which, combined with the computed enclosed mass of the
potential, $m_{enc}$, sets the instantaneous tidal radius and escape
velocity,
\begin{equation}
  R_{tide}=\Big(\frac{m_{sat}}{3m_{enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{esc}=\sqrt{\frac{2Gm_{sat}}{R_{tide}}}.
\end{equation}

Qualitatively, when $D_{ps}=\sqrt{|\bs{q}_i|^2+|\bs{p}_i|^2} <
\sqrt{2}$, the star is probably recaptured by the satellite, however
this is not imposed as a hard constraint. Instead, we construct a 6D
distribution of the minimum phase-space vectors for each star over the
individual, integrated orbits. For each star, $i$, the phase-space
distance, $D_{ps}$, is computed at each timestep $t_j$, and the vector
with the minimum phase-space distance is stored
\begin{align}
  t^*_i &= \argmin_{t} D_{ps,i}\\
  A_{ik} &= (\bs{q}_i(t^*_i),\bs{p}_i(t^*_i)) % A_ik are really samples from a distribution, or we treat them as such
\end{align}
where $k\in\{1,6\}$. The variance of this distribution will be larger
for orbits integrated in an incorrect potential relative to the
distribution computed from the `true' orbital history of the
stars. Thus, the \emph{generalized variance} of the distribution,
computed for a given set of potential parameters, $\bs{\theta}$, is a
natural choice for the scalar objective function used in constraining
the potential of the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik}))
\end{equation}

>>>>

% --------------------------------------------------------------------
% Above here is from KVJ. Below is APW.
%

\section{Context and motivation}
\subsection{2\% distance errors for RR Lyrae in the halo}

\subsection{Gaia and the age of astrometry}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which, amongst a myriad of other
science goals, aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission (cite?), Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full, six-dimensional phase-space
maps of the Galaxy for heliocentric distances of up to \sim 10~kpc for
xxx stars.

\section{Method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_i, \bs{v}_i)|_{t=0}$, the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$\footnote{Future work will
  investigate a more general ... basis function blah blah.}. For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for some time (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates, $(\bs{q}_i(t_j), \bs{p}_i(t_j))$, are
computed using the instantaneous tidal radius, escape velocity, and
phase-space position of the progenitor. At a particular timestep,
these coordinates are defined as
\begin{equation}
  \bs{q}_i = \frac{\bs{x}_i -
    \bs{x}_p}{R_{tide}}\,\,\,\,\,\,\,\,\bs{p}_i = \frac{\bs{v}_i -
    \bs{v}_p}{v_{esc}}.
\end{equation}
These definitions require an estimate of the mass of the satellite,
$m_{sat}$, which, combined with the computed enclosed mass of the
potential, $m_{enc}$, sets the instantaneous tidal radius and escape
velocity,
\begin{equation}
  R_{tide}=\Big(\frac{m_{sat}}{3m_{enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{esc}=\sqrt{\frac{2Gm_{sat}}{R_{tide}}}.
\end{equation}

Qualitatively, when $D_{ps}=\sqrt{|\bs{q}_i|^2+|\bs{p}_i|^2} <
\sqrt{2}$, the star is probably recaptured by the satellite, however,
this is not imposed as a hard constraint. Instead, we construct a 6D
distribution of the minimum phase-space vectors for each star over the
individual, integrated orbits. For each star, $i$, the phase-space
distance, $D_{ps}$, is computed at each timestep $t_j$, and the vector
with the minimum phase-space distance is stored
\begin{align}
  t^*_i &= \argmin_{t} D_{ps,i}\\
  A_{ik} &= (\bs{q}_i(t^*_i),\bs{p}_i(t^*_i)) % A_ik are really samples from a distribution, or we treat them as such
\end{align}
where $k\in\{1,6\}$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars. Thus, the \emph{generalized variance} of the
distribution, computed for a given set of potential parameters,
$\bs{\theta}$, is a natural choice for the scalar objective function
used in constraining the potential of the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
